Fully Automatic Loop-free Program Synthesis by Bounded Model Checking

### 2013-10-04 paper/paper.tex
nudges sec 2

Our task is to find a program which satisfies some specifcation.  We can formalize this
notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
are then relations and programs $P$ are computable functions:

$$ \sigma \subseteq I \times O $$
$$ P : I \rightarrow O$$

The synthesis problem is then to determine the validity of the following formula, and
to find a witness $P$ if the following second-order formula is valid:

$$\exists P . \forall x \in I . \sigma(x, P(x))$$

Depending on the logic needed to express this formula the synthesis problem may
be decidable, semi-decidable or undecidable.  For many interesting logics, checking
the validity of a second order formula, such as the synthesis formula, is undecidable.
Even for logics in which the problem is decidable, the quantifier alternation means that
we do not have efficient decision procedures to verify our formula, although efficient
methods may exist for solving problems with only one quantifier.  In particular
if we are working in the logic of propositional satisfiability, we have SAT solvers
which are efficient-in-practice at solving queries with a single quantifier.
The corresponding problem with one level of quantifier alternation (2-QBF) is
complete for $NP^{NP}$, and current 2-QBF solvers are much less efficient than SAT solvers.

@@

\begin{figure}
\centering
\begin{minipage}{0.9\textwidth}
 \begin{lstlisting}[mathescape,language=C]
function synth(inputs) {
  $(i_1, \ldots, i_N)$ = inputs
  query := $\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
  result := decide(query)

@@

To synthesize tricky bitvector programs like this, it is natural for us to work
in the logic of quantifer free propositional formulae and to use a SAT solver as
the decision procedure.  However, we propose a slightly different tack, which is
to use a decidable fragment of C as our underlying logic.  The fragment we use
has the following restrictions:

@@

Other than this any C constructs can be used, which still leaves a very expressive
language with the nice property that safety is decidable using a single query to a
bounded model checker.  We will refer to this fragment as C-.  To instantiate the
abstract synthesis algorithm in C- we must express $I, O, \sigma$ and $P$ in C-,
then ensure that we can express the validity of the synthesis formula as a safety
property of the resulting C- program.

Our encoding is the following:

\begin{itemize}
 \item $I$ is the type \verb|int[N]|.
 \item $O$ is the type \verb|int[M]|.
 \item $\sigma$ is a function with signature
 \verb|int check(int in[N], int out[M])|. This function is the only component supplied
 by the user.
 \item $P$ is written in a simple RISC language $\mathcal{L}$.  Programs in $\mathcal{L}$
 have the type \verb|prog_t|.
 \item We supply an interpreter for $\mathcal{L}$ which is written in C-.  The signature
 of this interpreter is \\
 \verb|void exec(prog_t p, int in[N], int out[M])|.
\end{itemize}

>>>>

Our task is to find a program which satisfies some specifcation.  We can formalize this
notion as follows: fix an input set $I$ and an output set $O$.  Specifications $\sigma$
relate inputs and outputs, and programs $P$ map an input to an output:
%
$$ \sigma \subseteq I \times O $$
$$ P : I \rightarrow O$$

The synthesis problem is then to determine the validity of the following
second-order formula, and if so, to find a a suitable witness program $P$:
%
$$\exists P .\, \forall x \in I.\, \sigma(x, P(x))$$

Depending on the logic needed to express this formula the synthesis problem
may be decidable, semi-decidable or undecidable.  For many interesting
logics, checking the validity of a second-order formula is undecidable. 
Even for logics in which the problem is decidable, the quantifier
alternation means that we may not have an efficient decision procedure for
the formula above, although efficient methods may exist for solving problems
with only one quantifier.  In particular, in the case of propositional
satisfiability, we have SAT solvers which are efficient-in-practice at
solving queries with a single quantifier.  The corresponding problem with
one level of quantifier alternation (2-QBF) is complete for $NP^{NP}$, and
current 2-QBF solvers are much less efficient than SAT solvers.

@@

\begin{figure}
\centering
\begin{minipage}{0.9\textwidth}
 \begin{lstlisting}[mathescape,language=C,morekeywords={function}]
function synth(inputs) {
  $(i_1, \ldots, i_N)$ = inputs
  query := $\exists P . \sigma(i_1, P(i_1)) \land \ldots \land \sigma(i_N, P(i_N))$
  result := decide(query)

@@

To synthesize tricky bitvector programs like this, it is natural for us to work
in the logic of quantifer-free propositional formulae and to use a SAT solver as
the decision procedure.  However, we propose a slightly different tack, which is
to use a decidable fragment of C as our underlying logic.  The fragment we use
has the following restrictions:

@@

Any other C constructs are permitted, which still leaves a very expressive
language with the nice property that safety is decidable using a single query to a
bounded model checker.  We will refer to this fragment as C-.  To instantiate the
abstract synthesis algorithm in C- we must express $I, O, \sigma$ and $P$ in C-,
then ensure that we can express the validity of the synthesis formula as a safety
property of the resulting C- program.

Our encoding is the following:
%
\begin{itemize}
 \item $I$ is the type \verb|int[N]|.
 \item $O$ is the type \verb|int[M]|.
 \item $\sigma$ is a function with signature
 \verb|int check(int in[N], int out[M])|. This function is the only component supplied
 by the user.
 \item $P$ is written in a simple RISC-like language $\mathcal{L}$.  Programs in $\mathcal{L}$
 have the type \verb|prog_t|.
 \item We supply an interpreter for $\mathcal{L}$ which is written in C-.  The signature
 of this interpreter is \\
 \verb|void exec(prog_t p, int in[N], int out[M])|.
\end{itemize}
